---
title: "Forestfires"
output:
  word_document: default
  pdf_document: default
  html_document: default
date: "2024-04-28"
---

Install packages

```{r}
install.packages("readxl")
library(readxl)
library(dplyr)
library(ggplot2)
```

Install BBC packages

```{r}
if(!require(pacman))install.packages("pacman")

pacman::p_load('dplyr', 'tidyr', 'gapminder',
               'ggplot2',  'ggalt',
               'forcats', 'R.utils', 'png', 
               'grid', 'ggpubr', 'scales',
               'bbplot')
```

```{r}
install.packages('rsconnect')
```

```{r}
devtools::install_github('bbc/bbplot')
```

Read the Excel files of forestfires from Jan 1, 2020 to April 28, 2024

```{r}


data_2020 <- read_excel("ForestFire_Nepal_aJan_April_20.xls")
data_2020_21 <- read_excel("ForestFire_Nepal_April_20_21.xls")
data_2021_22 <- read_excel("ForestFire_Nepal_April_21_22.xls")
data_2022_23 <- read_excel("ForestFire_Nepal_April_22_23.xls")
data_2023_24 <- read_excel("ForestFire_Nepal_April_23_24.xls")
```

Merge multiple data frames

```{r}
data_2020_24 <- rbind(data_2020, data_2020_21, data_2021_22, data_2022_23, data_2023_24)
```

Filter data with confidence \>70 only (f0)

```{r}
data_2020_24_f0 <- subset(data_2020_24, Confidence > 70)
```

Clean data by removing similar rows (f1)

```{r}
data_2020_24_f1 <- distinct(data_2020_24_f0)
```

Clean data by removing columns with missing values 'protected areas' (f2)

```{r}
data_2020_24_f2 <- data_2020_24_f1 %>% select(which(colSums(is.na(data_2020_24_f1)) == 0))
```

Extract the 'Year' and 'Month' from the "Date" column

```{r}
data_2020_24_f2$Date <- as.Date(data_2020_24_f2$Date)
data_2020_24_f2$Year <- format(as.Date(data_2020_24_f2$Date), "%Y")
data_2020_24_f2$Month <- format(as.Date(data_2020_24_f2$Date), "%m")

```

Calculate the count of fire incidents for each year

```{r}
yearly_fire_counts <- table(data_2020_24_f2$Year)
```

Convert the counts to a data frame

```{r}
fire_counts_df <- as.data.frame(yearly_fire_counts)
names(fire_counts_df) <- c("Year", "Fire_Count")
```

Plot the bar graph

```{r}
plot_bar <- ggplot(fire_counts_df, aes(x = Year, y = Fire_Count)) +
  geom_bar(stat = "identity", fill = "#F46203", width = 0.5) +
  
  #adding in-bar labels
  geom_label(aes(x = Year, y = Fire_Count, label = round(Fire_Count, 0)),
             hjust = 0.5, 
             vjust = 1, 
             colour = "white", 
             fill = NA, 
             label.size = NA, 
             family="Helvetica", 
             size = 4) +

   geom_hline(yintercept = 0, size = 1, colour="#333333") +
   bbc_style() +
  labs(title = "Forest Fires in Nepal on the rise again", subtitle = "Forest fire incidents surged by over 100 % per year since the 2021 record", x=NULL, y=NULL)+
  theme(
    plot.title = element_text(size = 20),  
    plot.subtitle = element_text(size = 14, margin=margin(b=50))
  )


#Finalize the plot

finalise_plot(plot_name = plot_bar,
              source = "Source: MoFE, Nepal. 2024 figures are until April 28.",
              save_filepath = "ff0.png",
              width_pixels = 640,
              height_pixels = 450,
              )


```

#Group data by year and month, calculate sum of fire counts

```{r}
monthly_fire_count <- data_2020_24_f2 %>%
  group_by(Year, Month, .groups = "drop") %>%
  summarise(Fire_Count = n())

# Create a data frame with complete sequences of years and months
complete_dates <- expand.grid(
  Year = unique(monthly_fire_count$Year),
  Month = sprintf("%02d", 1:12)  # Format month with leading zero
)

# Convert Year and Month to character type
complete_dates$Year <- as.character(complete_dates$Year)
complete_dates$Month <- as.character(complete_dates$Month)

# Perform left join to merge with existing data frame
complete_df <- left_join(complete_dates, monthly_fire_count, by = c("Year", "Month"))

# Replace missing Fire_Count values with 0
complete_df$Fire_Count[is.na(complete_df$Fire_Count)] <- 0

# Convert Fire_Count to integer
complete_df$Fire_Count <- as.integer(complete_df$Fire_Count)

# Change the month format to abbreviated month names
complete_df$Month <- month.abb[as.integer(complete_df$Month)]

# Convert the "Month" column to a factor with ordered levels from Jan to Dec
complete_df$Month <- factor(complete_df$Month, levels = month.abb, ordered = TRUE)


# Now complete_df contains all months for each year with Fire_Count values filled in,
# including missing months with Fire_Count as 0

```

#Plot the multiple line chart

```{r}

  
#Plot the graph
plot_line <- ggplot(complete_df, aes(x = Month, y = Fire_Count, group = Year, colour = as.factor(Year))) +
  geom_line(size = 1) +
  ylim(c(0,1400)) +
  geom_point(size = 1, shape = 5, fill = "white") +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  scale_colour_manual(values = c("#0078FF", "#1380A1", "#C76ADA", "#E4134F", "#F46203")) + 
  bbc_style() +
  theme(legend.position = "bottom") +
  labs(title="Forest Fires in Nepal on the rise again",
       subtitle = "'Dry and Windy' April month accounts for the most forest fires in the last five years") +
theme(
    plot.title = element_text(size = 22),  
    plot.subtitle = element_text(size = 14)  
  )

#Finalize the plot

finalise_plot(plot_name = plot_line,
              source = "Source: MoFE, Nepal. 2024 figures are until April 28.",
              save_filepath = "ff1.png",
              width_pixels = 640,
              height_pixels = 450,
              )

  
```

Line Chart:

```{r}
ggplot(fire_counts_df, aes(x = Year, y = Fire_Count,  group = 1)) +
    geom_line(colour = "#F46203", size = 1) +
  ylim(c(0,2000)) +
    geom_hline(yintercept = 0, size = 1, colour="#333333") +
    bbc_style() +
  labs(title="Forest Fires in Nepal",
       subtitle = "Number of forest fires reported in last five years") + 
   theme(
  axis.ticks.x = element_line(colour = "#333333"), 
  axis.ticks.length =  unit(0.26, "cm"))
```
